---
title: "From .mat to on-task: Complete Eye-Tracking Pipeline"
author: "Ari Dyckovsky, Peter Sokol-Hessner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From .mat to on-task: Complete Eye-Tracking Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

This vignette walks through the complete `itrackvalr` pipeline, from ingesting MATLAB `.mat` files through on-task classification. We'll demonstrate both **eager (tibble)** and **lazy (polars LazyFrame)** processing approaches.

## The CSN Study

The Clock-Signal-Noise (CSN) study uses the **EyeLink 1000 Plus** eye tracker in **remote mode**:

- **Tracking**: Monocular right eye, no chinrest
- **Sampling**: 500-1000 Hz
- **Remote mode**: Target sticker on forehead for distance measurement[^1]
- **Lens**: 25mm remote lens (better quality than 16mm standard)

[^1]: EL1000Plus User Manual, Section 3.2. Remote mode uses an infrared-reflective target sticker placed on the participant's forehead for head distance estimation, enabling tracking without a chinrest.

## Step 1: Read .mat File

```{r read_data, eval=FALSE}
library(itrackvalr)

# Read synthetic participant data
mat_file <- system.file("extdata/synthetic/synthetic_01.mat", 
                         package = "itrackvalr")
data <- read_mat_data(mat_file)

# Inspect structure
str(data, max.level = 1)
```

Returns 4 tibbles:
- **samples**: High-frequency gaze data (30,000 rows)
- **events**: Discrete occurrences (132 rows)
- **behavioral**: Trial-level outcomes (60 rows)
- **metadata**: Session information (1 row)

## Step 2: Parse Validation Messages

```{r validation, eval=FALSE}
# Extract pre/post calibration metrics
validation_df <- parse_validation_msgs(data$events)

# View validation errors
validation_df
#> id    time  type  avg_err_deg  max_err_deg  offset_x_px  offset_y_px
#> SYN01 0     pre   0.64         1.28         12.3         8.5
#> SYN01 61000 post  0.68         1.32         14.1         9.2
```

## Step 3: Apply Calibration Offsets

### Eager Approach (Small Datasets)

```{r calibration_eager, eval=FALSE}
# Linear interpolation from pre to post offsets
samples_calibrated <- apply_calibration_offsets(
  data$samples,
  validation_df,
  method = "linear"
)

# New columns added:
# - x_px_adj, y_px_adj: Calibration-corrected coordinates
# - offset_x_applied, offset_y_applied: Applied offsets
```

### Lazy Approach (Large Datasets)

```{r calibration_lazy, eval=FALSE}
# Convert to LazyFrame for efficient processing
samples_lf <- to_polars_lf(data$samples)

# Apply calibration (remains lazy)
samples_calibrated_lf <- apply_calibration_offsets(
  samples_lf,
  validation_df,
  method = "linear"
)

# Data not yet loaded into memory!
# Stream to Parquet:
sink_parquet(samples_calibrated_lf, "output/calibrated/samples.parquet")
```

## Step 4: Resample to Uniform Grid

```{r resample, eval=FALSE}
# Standardize to 500 Hz with gap detection
samples_resampled <- resample_samples(
  samples_calibrated,
  hz_standard = 500,
  gap_threshold = 20  # ms
)

# New columns:
# - gap_flag: TRUE if interpolated across large gap
# - blink: TRUE if pupil missing/zero
```

## Step 5: Compute Clock Hand Dynamics

```{r clock_hand, eval=FALSE}
# Derive hand tip positions from task parameters
hand_positions <- derive_hand_dynamics(
  metadata_df = data$metadata,
  events_df = data$events,
  cfg = config::get(),
  sampling_rate_hz = 500
)

# Join to gaze samples
samples_with_hand <- join_hand_positions(
  samples_resampled,
  hand_positions
)

# New columns:
# - tip_x_px, tip_y_px: Clock hand tip position
# - angle_deg: Current angle (0 = 12 o'clock)
# - center_x_px, center_y_px: Clock center
```

## Step 6: Compute Zone of Uncertainty

```{r zone, eval=FALSE}
# Calculate acceptable distance threshold from validation errors
zone_radii <- compute_zone_radius(
  validation_df,
  strategy = "average",  # or "maximum", "pre_avg", "post_avg"
  scaling = 1.0
)

# Returns per-participant radii (typically 20-50 pixels)
zone_radii
#> id    r_deg  r_px   strategy  scaling
#> SYN01 0.66   23.6   average   1.0
```

## Step 7: Distance & On-Task Classification

```{r binarize, eval=FALSE}
# Compute Euclidean distance to hand tip
samples_with_distance <- compute_distance(
  samples_with_hand,
  use_adjusted = TRUE  # Use calibration-corrected coordinates
)

# Classify as on-task (within zone) or off-task
samples_binarized <- binarize_on_task(
  samples_with_distance,
  zone_radii
)

# New column:
# - on_task: TRUE (within zone), FALSE (outside), NA (blink/missing)

# Check overall proportion
mean(samples_binarized$on_task, na.rm = TRUE)
#> [1] 0.006  # 0.6% on-task (low for stochastic gaze)
```

## Step 8: Segment by Trials

```{r segment, eval=FALSE}
# Slice gaze data by image presentation trials
trial_segments <- segment_trials(
  samples_binarized,
  data$events,
  data$behavioral
)

# New columns:
# - trial: Trial number
# - t_trial_ms: Time since image onset (0 = onset)
# - t_absolute: Original timestamp
# - image_index: Which image was shown
# - outcome: hit, miss, false_alarm, correct_rejection
```

## Complete Pipeline Example

### Small Scale (≤10 participants): Eager Processing

```{r pipeline_eager, eval=FALSE}
# Read and process 5 synthetic participants
mat_files <- get_mat_files("inst/extdata/synthetic")
all_data <- read_all_participants(mat_files)

# Aggregate (auto-uses dplyr for small scale)
all_samples <- aggregate_samples(all_data)  # Returns tibble
all_validation <- aggregate_validation(all_data)
all_behavioral <- aggregate_behavioral(all_data)
all_metadata <- aggregate_metadata(all_data)

# Process eagerly
validation_summary <- summarize_validation_relationships(all_validation)
zone_radii <- compute_zone_radius(all_validation, strategy = "average")

samples_processed <- all_samples |>
  apply_calibration_offsets(all_validation, method = "linear") |>
  resample_samples(hz_standard = 500)

# Continue with hand dynamics, distance, binarization...
```

### Large Scale (>10 participants): Lazy Processing

```{r pipeline_lazy, eval=FALSE}
# Read all 57 participants
mat_files <- get_mat_files("/Volumes/shlab/Projects/CSN/data")
all_data <- read_all_participants(mat_files)

# Aggregate (auto-uses polars for large scale)
all_samples_lf <- aggregate_samples(all_data)  # Returns LazyFrame!

# Process lazily (no data loaded yet)
samples_processed_lf <- all_samples_lf |>
  apply_calibration_offsets(all_validation, method = "linear") |>
  resample_samples(hz_standard = 500) |>
  # join hand positions, compute distance, binarize...

# Stream to disk without loading into memory
sink_parquet(
  samples_processed_lf, 
  "output/preprocessed/ALL_samples_binarized.parquet"
)

# Memory usage: <2 GB vs >10 GB with eager approach!
```

## Using the targets Pipeline

The recommended approach is the `targets` pipeline:

```{r targets_pipeline, eval=FALSE}
library(targets)

# Run complete pipeline
tar_make()

# Visualize dependencies
tar_visnetwork()

# Inspect specific targets
tar_read(samples_binarized)
tar_read(cohort_summary)
```

## Performance Comparison

| Scale | Approach | Time | Memory | Notes |
|-------|----------|------|--------|-------|
| 5 participants | dplyr (eager) | 1 min | <3 GB | Fast enough |
| 57 participants | dplyr (eager) | ~7 min | >10 GB | Memory limited |
| 57 participants | polars (lazy) | <1 min | <2 GB | **Recommended** |

The package automatically chooses the optimal backend via smart dispatch!

## References

1. [EyeLink 1000 Plus User Manual](https://www.hse.ru/mirror/pubs/share/560338728.pdf)
2. [tidypolars documentation](https://tidypolars.etiennebacher.com/)
3. [itrackvalr GitHub](https://github.com/aridyckovsky/itrackvalr)

## Session Info

```{r session_info, eval=FALSE}
sessionInfo()
```


